---
title: "Presidential Election Maps for United States from 1920s"
author: "Team FiveThirtyNine"
date: "12/13/2017"
output: rmarkdown::html_document
runtime: shiny
---
# Introduction

Our goal for this project was to create an interactive electoral map displaying results of U.S. presidential elections from 1920 to present. With this aim in mind, we create a visualization similar to other maps summarizing results from the Electoral College. Using leaflet within a Shiny app, we enable the user to interactively update the electoral map as well as explore the state level results for a given election. We also give users access to the data used to create the map for a particular election year which includes results for all candidates across all states.

```{r, message=FALSE, echo=FALSE, warning=FALSE}
library(stringr)
library(dplyr)
library(tidyr)
library(shiny)
library(rvest)
library(sf)
library(leaflet)
devtools::install_github("hrbrmstr/albersusa")
library(albersusa)
```

# Scraping Data from Wikipedia

While there are numerous data sources available summarizing presidential election results, none are as complete as the election results tables on Wikipedia. Unlike other sources, these tables include state by state results for all presidential candidates, not just those from the two major parties along with a subset of third party candidates. Therefore, using Wikipedia is not only convenient as it allows us to obtain data from a single source for all election years, but it also gives us more refined data on the performance of all third party candidates. We create a function `get_election_data(year)` to extract data on election results for the year specified in its argument.

This function generates a clean data table for each election year via scraping data from Wikipedia's presidential election pages. Because there is usually more than one table on each Wikipedia page, we filter the desired table by measuring the dimension of table. The table with more than 50 rows is supposed to contain election data by state, and thus we pick that table for each election year. We use regular expressions to clean the scraped table (i.e. removing all extraneous symbols), as well as to standardize column names. Additionally, we remove superfluous columns/rows, add a column that indicates the number electoral votes for each state if such a column does not already exist, mutate the `State` column to contain state abbreviations, add a column that indicates winner for each state, and change negative values (such as the margin of victory as a percent) to their absolute values. All of this data cleaning is done so that the scraped tables are uniform across all years. 

For some election years, there is more than one row corresponding to Maine and Nebraska since these states allocate electoral votes according to results in congressional districts as well as the the overall state popular vote. These electoral rules were adopted in these states beginning in the 1972 and 1992 elections, respectively. There are election years after these years in which the results are not split by district for either Maine or Nebraska. However, there is no uniform criterion detailing when the Wikipedia table splits results by district for each of these two states. Therefore, if the data for these states is split by district we keep the row corresponding to the overall state popular vote, and discard the rows detailing election results in the congressional districts. In doing so we sacrifice a minimal amount of information so that the election data tables are consistent over all years.

```{r}
get_election_data=function(year){

  html_str<-paste0("https://en.wikipedia.org/wiki/United_States_presidential_election,_",year,"#Results_by_state")
 
  
  wiki<-read_html(html_str)
  wikitables<-wiki%>%html_nodes(".wikitable.sortable")
  
  for(i in 1:length(wikitables)){
    temp=wikitables[[i]]%>%html_table(fill=TRUE)
    if(nrow(temp)>49)
      {
        result_table=temp
        break
      }
  }
  
  #set uniform column names
  if(year == 1992){
    colnames(result_table)<-paste(result_table[1,],result_table[2,]) %>%
      str_replace_all("\n"," ") %>%
      str_replace_all("^\\s","")
    result_table<-result_table[-c(1,2),]
  }
  else{
    colnames(result_table)<-paste(colnames(result_table),result_table[1,]) %>%
      str_replace_all("\n"," ") %>%
      str_replace_all("^\\s","")
    result_table<-result_table[-1,]
  }
  
  d = result_table
  
  colnames(d) = colnames(d) %>%
    str_replace_all("\\s$","") %>%
    str_replace_all("^E$","electoral votes") %>%
    str_replace_all("(?<!^)[Ee]lectoral votes","E") %>%
    str_replace_all("Top.+/R\\)","Margin") %>%
    str_replace_all("#","Vote") %>%
    str_replace("State [Tt]otal Vote","Total Vote") %>%
    str_replace("State [Tt]otal$|Total State$|^Total$","State Abbr") %>%
    str_replace("State or district","State") %>%
    str_replace_all("\\★","")
  
  #remove superfluous columns
  omit <- c("Electoral method","Total Status")
  d = d[,which(!(colnames(d) %in% omit))]
  
  #formatting dataframe
  dd = apply(d,2,function(x){
    x %>% str_replace_all("–|−","-") %>% str_replace_all("N/A","-") %>%
      str_replace_all("^-$|^$","0") %>% str_replace_all("%","") %>%
      str_replace_all(",","") %>% str_replace_all("\\†","") %>%
      str_replace_all("\\★","")
    }) %>% as.data.frame(stringsAsFactors = FALSE)
  
  #turn appropriate columns into numeric variables
  idx = which(str_detect(colnames(dd),"State"))
  
  dd[,-idx] = lapply(dd[,-idx],as.numeric)
  dd$`Margin Vote`<-abs(dd$`Margin Vote`)
  dd$`Margin %`<-abs(dd$`Margin %`)
  
  
  #add electoral votes column if doesn't already exist
  if(!("electoral votes" %in% colnames(dd))){
    dd = dd %>% mutate(
      elec_vote = rowSums(dd[,str_detect(colnames(dd),"E$")])
    )
    colnames(dd)[colnames(dd) == "elec_vote"] = "electoral votes"
  }
  
  #add winner column and alter state column
  dd = dd %>%
    mutate(Winner = apply(dd,1,function(x){names(
      which.max(x[str_detect(colnames(dd),"E$")])) %>% 
        str_replace("\\sE$","")}),
      State = `State Abbr` %>% str_extract_all("[:upper:]{2}") %>% unlist()
      ) %>% select(-`State Abbr`)
  
  #alter table to contain distinct rows (Maine & Nebraska)
  clean_table = dd %>%distinct(State,.keep_all = TRUE)

  return(clean_table)
}
```


# Shiny App

We visualize our election data via utilizing leaflet within a Shiny app. We create two tab panels for this app: the first shows the electoral map generated using spatial data from the "albersusa" library while the second outputs a table of election results filtered based on the user's selection. Furthermore, the first panel includes pie charts for the popular and electoral vote as part of the visualization and allows users to see state by state results via generating pop-ups when a specific state is clicked.

In the first panel, when an election year is entered and the action button is clicked, the server calls the `get_election_data(year)` function, which returns a tidy dataframe containing the election results. In order to update the map, we create a new column called "Party" to indicate the party affiliation for the winner of each state, and assign different colors to our visualization map according to this column (blue=democratic, red=republican; green=third party; white=state not part of the Electoral College in a given year). When a candidate from one of the two major parties wins, we take into account the margin of victory in a particular state by assigning lighter colors to states where the margin of victory was small (less than 10 percentage points). We then merge a subset of the columns of the dataframe generated by `get_election_data(year)` along with `Party` to the dataframe containing the geometry objects for each state supplied in "albersusa". The data from this object is then used to update the electoral map. 

Pie charts are created by summing up electoral/popular votes from the dataframe generated by `get_election_data(year)`. Pop-ups containing state results are created using a method inspired by the SuperZip app from the Shiny gallery. We create a funciton called `pop_content()`, which outputs a list of election results corresponding to the state clicked by the user (current state, winner, margin of victory (%), etc.). When a click is observed, the server calls this function using the name, latitude, and longitude associated with the state clicked by the user as inputs.

For the second panel, the choices in selection boxes are generated when the a year is entered by user and the action button is clicked in the first panel. The user can then filter the dataframe containing the election results for that year by candidate and by state upon clicking the action button "See Candidate Data". We note that the user must specify an option in at least one of the selection boxes for the desired dataframe to render.

```{r}
#map using albersusa
usa <- usa_sf() %>% mutate(State = as.character(iso_3166_2))
pal <- colorNumeric(c("white","red","indianred2",
                      "cornflowerblue","blue","green"),0:5)

shinyApp(
  ui = navbarPage("Electoral College",
    tabPanel("Electoral Map",
        div(class="outer"),
        leafletOutput("map", width="100%",height=600),
        absolutePanel(id = "elec_dat",top = 70,
                   left = 50,right = "auto", bottom = "auto",
                   width = 300, height = "auto", draggable = TRUE, fixed = TRUE,
          h3("Presidential Election App"),
          numericInput("yearr","Year", min = 1920, max = 2016,
                       step = 4, value = 1996),
          actionButton("do", "See Results"),
          checkboxInput("electoral","Display Electoral Votes?",value=FALSE),
          plotOutput("pop_vote"),
          dataTableOutput("elec_vote"),
          style = "opacity: 0.75"
        )
    ),
    tabPanel("Election Data",
        sidebarLayout(
          sidebarPanel(
            selectInput("cand","Candidate",choices = NULL),
            selectInput("state","State", choices = NULL),
            actionButton("retrieve", "See Candidate Data")
          ),
        mainPanel(
          dataTableOutput("elec_dat")
        )
      )
    )
  ),
  server = function(input, output, session) 
  {
    
    #reactive values
    data <- eventReactive(input$do,{
      get_election_data(input$yearr)
    })
    
#Panel 1
    #Categorize winning conditions for each state (strong win, weak win, etc)
    k <- reactive({
      data() %>%
        mutate(`Party` = case_when(
          str_detect(.$`Winner`,"Republican") & .$`Margin %`>10 ~ 1,
          str_detect(.$`Winner`,"Republican") & .$`Margin %`<=10 ~ 2,
          str_detect(.$`Winner`,"Democratic") & .$`Margin %`<=10 ~ 3,
          str_detect(.$`Winner`,"Democratic") & .$`Margin %`>10 ~ 4,
          TRUE ~ 5)) %>%
        select(State,`Winner`,`electoral votes`,`Margin %`,`Party`)
    })
    
    #add some columns to albersusa dataframe
    kk <- reactive({
      left_join(usa,k(), by ="State") %>%
        mutate(`Party` = ifelse(is.na(.$`Party`),0,.$`Party`),
               `electoral votes` = ifelse(
                 is.na(.$`electoral votes`),0,.$`electoral votes`),
               `Margin %` = ifelse(is.na(.$`Margin %`),0,.$`Margin %`))
    })
    

    
    #render pie chart for popular vote/electoral vote
    output$pop_vote <- renderPlot({
       r_lab = colnames(data()) %>% str_extract("\\w+(?=\\sRepublican)") %>%
                   na.omit() %>% unique()
       d_lab = colnames(data()) %>% str_extract("\\w+(?=\\sDemocratic)") %>%
                   na.omit() %>% unique()
      
      if(input$electoral=="FALSE"){
       pi_r = data()[nrow(data()),which(str_detect(
         colnames(data()),"Republican Vote"))[1]]
       pi_d = data()[nrow(data()),which(str_detect(
          colnames(data()),"Democratic Vote"))[1]]
        pi_o = data()[nrow(data()),which(str_detect(
          colnames(data()),"Total Vote"))[1]] - sum(pi_r,pi_d)
        x = c(pi_r,pi_d,pi_o)
      
       pie(x, labels = paste0(round(100*x/sum(x), 1),"%"),
            main = "Popular Vote",col=c("red","blue","grey"))
       legend("topleft",
              c(r_lab[1],d_lab[1],"Others"),
              fill = c("red","blue","grey"), cex = 0.75)
       
      }else{
        pi_r = data()[nrow(data()),which(str_detect(
         colnames(data()),"Republican E"))[1]]
       pi_d = data()[nrow(data()),which(str_detect(
          colnames(data()),"Democratic E"))[1]]
        pi_o = data()[nrow(data()),which(str_detect(
          colnames(data()),"electoral votes"))[1]] - sum(pi_r,pi_d)
        x = c(pi_r,pi_d,pi_o)
        
       
       pie(x, labels = x,
            main = "Electoral Vote",col=c("red","blue","grey"))
       legend("topleft",
              c(r_lab[1],d_lab[1],"Others"),
              fill = c("red","blue","grey"), cex = 0.75)
      }
    })

    
    #render intial map
    output$map <- renderLeaflet({
      leaf = leaflet(usa) %>%
        addPolygons(smoothFactor = 0.2, fillOpacity = 0.8,
                    fillColor = "gray",color = "black",
                    weight = 1, opacity = 1) %>% addLegend(position ="bottomright",
                  colors = c("red","pink","cornflowerblue","blue",
                          "green","white"),
                  labels = c("Strong Rep.","Weak Rep.","Weak Dem.","Strong Dem.",
                  "Other","No Info"),
                  title = "Party")
      })
    #update map with colors
    observe({
      leafletProxy("map", data = kk()) %>%
        addPolygons(smoothFactor = 0.2, fillOpacity = 0.8,
                    fillColor = ~pal(Party),color = "black",
                    weight = 1, opacity = 1, layerId =~name)
    })
    
    
    #function for popup content
    pop_content <- function(state,lat,lng){
      selectedState = kk() %>% filter(name == state)
      if(is.na(selectedState$`Winner`)){
        content = as.character(tagList(
          tags$h4(selectedState$name),
          sprintf("No Information Available")
        ))

      }else{
        content <-as.character(tagList(
          tags$h4(selectedState$name),
          sprintf("Electoral votes: %s", selectedState$`electoral votes`),tags$br(),
          sprintf("Winner: %s",selectedState$`Winner`),tags$br(),
          sprintf("Margin: %s%%",abs(selectedState$`Margin %`))
        ))
      }

      leafletProxy("map") %>% addPopups(lng, lat, content, layerId = state)
    }
    #when a click is observed, check if it is valid: if it is, then call the previous function to pop up contents.
    observe({
      leafletProxy("map") %>% clearPopups()
      event <- input$map_shape_click
      if(is.null(event)){
       return()
      }
      isolate({
        pop_content(event$id,event$lat,event$lng)
      })
    }) 

    
 #Panel 2   
    
    #update select choices for panel 2
    list_cand<-eventReactive(input$do,{
      colnames(data()) %>%
        str_extract("^(?!Other[s\\s]|Total|Margin)(.*?)[:upper:]\\w+(.*?)[:upper:]\\w{2,}")%>%
        na.omit()%>%unique()
    })
    observe({
      updateSelectInput(session, "cand","Candidate",
                      choices=c("none",list_cand()))
    })
    
    observe({
      updateSelectInput(session, "state","State",
                      choices=c("none",data()$State))
    })
    
    #output dataTable based on users selection for panel 2
    cand_table<-eventReactive(input$retrieve,{
       if(input$cand!="none"&&input$state!="none"){
        a<-data()%>%select(starts_with(input$cand))
        aa<-cbind(data()$State,a)
        colnames(aa)<-c("State","Popular Vote","Percentage","Electoral Vote")
        aa%>%filter(State==input$state)
       }else if(input$state=="none"&&input$cand!="none"){
         a<-data()%>%select(starts_with(input$cand))
         aa<-cbind(data()$State,a)
         colnames(aa)<-c("State","Popular Vote","Percentage","Electoral Vote")
         aa
       }else if(input$state!="none"&&input$cand=="none"){
         a<-data()%>%filter(State==input$state)
        a
       }else{
         NULL
       }
      
    })
    output$elec_dat<-renderDataTable(
      cand_table())
      
  }
  
)

```


